---
title: "Reproducible Research: Peer Assessment 1"
author: "David Kazaryan aka 1x0r"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      root.dir='/media/david/DATA/Dropbox/Work/git/learning/RepData_PeerAssessment1',
                      echo = TRUE)
```

## Loading and preprocessing the data

Frst of all, let's load a few libraries.

```{r, message = FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(mice)
```

The dataset is opened directly from the archive:

```{r}
f <- unz(file.path(getwd(), "activity.zip"), "activity.csv")
amd <- read.csv(f)
```

For the future steps we have to transform the "date" column into POSIXct object:

```{r}
amd$date <- parse_date_time(amd$date, c("%Y%-%m%-%d%"))
```

## What is mean total number of steps taken per day?
```{r}
total_steps <- tapply(amd$steps, amd$date, sum)
qplot(total_steps, geom = "histogram", bins = 25, na.rm = TRUE,
      fill = I("blue"), alpha = I(0.5), col = I("black"),
      xlab = "Steps", main = "Number of steps taken each day")
```

```{r}
mean_steps <- mean(total_steps, na.rm = TRUE)
median_steps <- median(total_steps, na.rm = TRUE)
```
The mean of the total number of steps taken per day is `r sprintf("%.2f", mean_steps)`.
The median of the total number of steps taken per day is `r median_steps`.

## What is the average daily activity pattern?

```{r}
steps_per_interval <- tapply(amd$steps, amd$interval, mean, na.rm = TRUE)
ggplot(data = as.data.frame(steps_per_interval), 
       aes(x = unique(amd$interval), y = steps_per_interval)) + 
  geom_line() +
  ggtitle("Average number of steps per time interval") +
  labs(x = "5-minute time interval",
       y = "Number of steps")
  
```
```{r}
max_steps <- which(steps_per_interval == max(steps_per_interval, na.rm = TRUE))
```

`r max_steps`th 5-minute interval contains the maximum number of steps.

## Imputing missing values
```{r}
missing_count <- sum(rowSums(is.na(amd)) > 0)
```

`r missing_count` rows with missing values are in the dataset

The only column tht contains missing values is *steps*. Let's impute them with rounded average number of steps in the corresponding 5-minutes interval:

```{r}
imp_amd <- amd %>% 
      group_by(interval) %>% 
      mutate(steps = replace(steps, is.na(steps), round(mean(steps, na.rm = TRUE)))) %>% 
      ungroup %>%
      as.data.frame
```

*imp_amd* is the new *data.frame* with imputed missing values. Let's plot a histogram with *imp_amd*.

```{r}
imp_total_steps <- tapply(imp_amd$steps, imp_amd$date, sum)
qplot(imp_total_steps, geom = "histogram", bins = 25, na.rm = TRUE,
      fill = I("blue"), alpha = I(0.5), col = I("black"),
      xlab = "Steps", main = "Number of steps taken each day with imputed NAs")
```

Histogram visibly differs from the histogram with missing values omitted.

```{r}
imp_mean_steps <- mean(imp_total_steps)
imp_median_steps <- median(imp_total_steps)
```

The mean of the total number of steps taken per day is `r sprintf("%.2f", imp_mean_steps)`.
The median of the total number of steps taken per day is `r sprintf("%.2f", imp_median_steps)`.

Mean and median values slightly changed from their previous values calculated from the dataset with ignored NAs.

## Are there differences in activity patterns between weekdays and weekends?

I introduced a new variable with two levels: weekends and weekdays. By default my Ubuntu has Russian language set up, so weekend abbreviations look like *Сб* for *S*, Saturday and *Вс* for *U*, Sunday.
```{r}
imp_amd$weekday <- factor(weekdays(imp_amd$date, abbreviate = TRUE) %in% c("Сб", "Вс"), 
                          levels = c(TRUE, FALSE), labels = c("weekend", "weekday"))

ggplot(imp_amd, aes(x = interval, y = steps)) + geom_point(size = 0.5, color = "gray") +
  facet_grid(. ~ weekday) +
  stat_summary(fun.y = mean, geom = "line", color = "red") +
  ggtitle("Average number of steps per time interval") +
  labs(x = "5-minute time interval",
       y = "Number of steps")
```

Person activity in weekends seem to start later than on weekdays (probably due to the lack of necessity in getting up to go to work or school).